---
title: "Transfer Analyses"
author: "Virginia Marchman and Janet Bang"
date: "April 15, 2019; updated September 20, 2019; Updated for R1 of paper January 21, 2020"
output:
  html_document: 
   toc: true
   toc_float: true
---

```{r, echo = F}
library(knitr)
opts_chunk$set(echo=TRUE, 
               warning=FALSE, message=FALSE, 
               cache=FALSE)
options(width = 100)
```

This is the code to generate the analyses in Marchman, Bermudez, Bang & Fernald (2019), Off to a good start: Early Spanish-language processing efficiency supports Spanish- and English-language outcomes at 4 ½ years in sequential bilinguals. Submitted Oct 2019; Revisions made January 2020; Final manuscript accepted for publication April 2020.

Load libraries and set theme
```{r}
library(tidyverse)
library(effsize)
library(stargazer)
library(psych)
library(powerAnalysis)

theme_set(theme_bw())
```


# Data prepping
### Load data
```{r}
transfer <- read_csv("transfer.csv") %>% 
  dplyr::select(-X1)
```



### Check data
```{r}
# Sex - 41 boys, 54 girls
transfer %>% 
  group_by(Sex) %>% 
  count()
```



### Transform variables (e.g., character to factor)
```{r}
transfer <- transfer %>% 
  mutate(ID_1 = factor(ID_1), 
         MomCountryBirth = factor(MomCountryBirth),
         Sex = factor(Sex))
```




# Demographic variables
### Descriptive statistics
```{r}
# All descriptives for continuous variables
age_descriptives <- sapply(transfer[, 4], describe)

other_descriptives <- sapply(transfer[, 9:22], describe)

cbind(age_descriptives, other_descriptives)
```



### Percentage of mothers by country of birth
```{r}
transfer %>% 
  group_by(MomCountryBirth) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(percentage = n/sum(n))
```


### Percentage of children by birth order
```{r}
transfer %>% 
  group_by(BirthOrder) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(percentage = n/sum(n))
```



### Number of trials for RT
```{r}
transfer %>% 
  summarise(mean_trials = mean(DRT25mKnownN, na.rm = T), 
            sd_trials = sd(DRT25mKnownN, na.rm = T),
            min_trials = min(DRT25mKnownN, na.rm = T), 
            max_trials = max(DRT25mKnownN, na.rm = T))
```



### Number of trials for Accuracy
```{r}
transfer %>% 
  summarise(mean_trials = mean(ACC25m3001800KnownN, na.rm = T), 
            sd_trials = sd(ACC25m3001800KnownN, na.rm = T),
            min_trials = min(ACC25m3001800KnownN, na.rm = T), 
            max_trials = max(ACC25m3001800KnownN, na.rm = T))
```



# Language Background at 2 and 4.5 years
```{r}
# Comparing LBQ OVERALL at 2 and 4.5 yrs
t.test(transfer$LBQSpan25m, transfer$LBQSpan_4.5yrs, paired = TRUE, alternative = "two.sided")
effsize::cohen.d(transfer$LBQSpan25m, transfer$LBQSpan_4.5yrs, paired = T)

# Comparing LBQ CAREGIVERS at 2 and 4.5 yrs
t.test(transfer$LBQSpan25mCaregivers, transfer$LBQSpanCaregivers_4.5yrs, paired = TRUE, alternative = "two.sided")
effsize::cohen.d(transfer$LBQSpan25mCaregivers, transfer$LBQSpanCaregivers_4.5yrs, paired = T)

# Correlations between LBQ OVERALL and LBQ CAREGIVERS
# At 2 years
cor.test(transfer$LBQSpan25m, transfer$LBQSpan25mCaregivers)
# At 4.5 years
cor.test(transfer$LBQSpanCaregivers_4.5yrs, transfer$LBQSpan_4.5yrs)

```



# Child language measures at 2 years
```{r}
# Accuracy - t-test against chance
t.test(transfer$GoodACC25m3001800Known, y = NULL, mu = .50, var.equal = FALSE)
(mean(transfer$GoodACC25m3001800Known) - .50) / sd(transfer$GoodACC25m3001800Known)

# Accuracy and RT
cor.test(transfer$GoodACC25m3001800Known, transfer$GoodDRT25mKnown)

# Correlations between CDI and Processing measures
# CDI and Accuracy
cor.test(transfer$CDIVocPost25, transfer$GoodACC25m3001800Known)

# CDI and RT
cor.test(transfer$CDIVocPost25, transfer$GoodDRT25mKnown)


```



# Child language measures at 4.5 years
```{r}
# Comparing Spanish and English composite scores
t.test(transfer$Spancomposite, transfer$Engcomposite, paired = TRUE, alternative = "two.sided")
effsize::cohen.d(transfer$Spancomposite, transfer$Engcomposite)

# Spanish composite scores against norm
t.test(transfer$Spancomposite, y = NULL, mu = 100, var.equal = FALSE)
ES.t.one(m=92.5,sd=17.3,mu=100)

# English composite scores against norm
t.test(transfer$Engcomposite, y = NULL, mu = 100, var.equal = FALSE)
ES.t.one(m=87.7,sd=13.1,mu=100)

# Correlation between Spanish and English composite scores
cor.test(transfer$Spancomposite, transfer$Engcomposite)

```




# Regression models
### Center covariates
```{r}
# The "c" function turns the class from matrix to numeric
# scale = divides by the standard deviation
transfer <- transfer %>% 
  mutate(age_4.5y_zscore = c(scale(Age, scale = T)), 
         hi_4.5y_zscore = c(scale(HI_4.5yrs, scale = T)), 
         lbq_4.5y_zscore = c(scale(LBQSpan_4.5yrs, scale = T)), 
         cdi_25m_zscore = c(scale(CDIVocPost25, scale = T)), 
         rt_zscore = c(scale(GoodDRT25mKnown, scale = T)),
         acc_zscore = c(scale(GoodACC25m3001800Known, scale = T)))
```



### Models - Spanish composite
```{r}
# covariates only

m1 <- lm(Spancomposite ~ hi_4.5y_zscore + lbq_4.5y_zscore, 
         data = transfer)

# covariates + cdi
m2 <- lm(Spancomposite ~ hi_4.5y_zscore + lbq_4.5y_zscore + cdi_25m_zscore, 
         data = transfer)

# covariates + acc
m3 <- lm(Spancomposite ~ hi_4.5y_zscore + lbq_4.5y_zscore + acc_zscore, 
         data = transfer)

# covariates + cdi + acc
m4 <- lm(Spancomposite ~ hi_4.5y_zscore + lbq_4.5y_zscore + cdi_25m_zscore +
           acc_zscore, 
         data = transfer)

# covariates + rt
m5 <- lm(Spancomposite ~ hi_4.5y_zscore + lbq_4.5y_zscore + rt_zscore, 
         data = transfer)

# covariates + cdi + rt
m6 <- lm(Spancomposite ~ hi_4.5y_zscore + lbq_4.5y_zscore + cdi_25m_zscore + 
           rt_zscore, 
         data = transfer)

anova(m1, m2)
anova(m1, m3)
anova(m1, m4)
anova(m1, m5)
anova(m1, m6)

stargazer(m1, m2, m3, m4, m5, m6, type = "text",
         star.char = c(".","*","**","***"),
         star.cutoffs = c(.1, .05, .01, .001),
         notes = c(". p<0.1; * p<0.05; ** p<0.01; *** p<0.001"),
         notes.append = F,
         digits = 3,
         dep.var.labels = c("Spanish Language Composite"),
         covariate.labels=c("SES 4.5y", "Spanish Language Exposure at 4.5y",
                            "Spanish Vocabulary at 25m",
                            "Spanish Accuracy at 25m",
                            "Spanish RT at 25m"))

```



### Models - English composite
```{r}
# covariates only
m7 <- lm(Engcomposite ~ hi_4.5y_zscore + lbq_4.5y_zscore, 
         data = transfer)

# covariates + cdi
m8 <- lm(Engcomposite ~ hi_4.5y_zscore + lbq_4.5y_zscore + cdi_25m_zscore, 
         data = transfer)

# covariates + acc
m9 <- lm(Engcomposite ~ hi_4.5y_zscore + lbq_4.5y_zscore + acc_zscore, 
         data = transfer)

# covariates + cdi + acc
m10 <- lm(Engcomposite ~ hi_4.5y_zscore + lbq_4.5y_zscore + cdi_25m_zscore
          + acc_zscore, 
         data = transfer)

# covariates + rt
m11 <- lm(Engcomposite ~ hi_4.5y_zscore + lbq_4.5y_zscore + rt_zscore, 
         data = transfer)

# covariates + cdi + rt
m12 <- lm(Engcomposite ~ hi_4.5y_zscore + lbq_4.5y_zscore + cdi_25m_zscore + 
           rt_zscore, 
         data = transfer)

anova(m7, m8)
anova(m7, m9)
anova(m7, m10)
anova(m7, m11)
anova(m7, m12)
stargazer(m7, m8, m9, m10, m11, m12, type = "text",
         star.char = c(".","*","**","***"),
         star.cutoffs = c(.1, .05, .01, .001),
         notes = c(". p<0.1; * p<0.05; ** p<0.01; *** p<0.001"),
         notes.append = F,
         digits = 3,
         dep.var.labels = c("English Language Composite"),
         covariate.labels=c("SES 4.5y","Spanish Language Exposure at 4.5y",
                            "Spanish Vocabulary at 25m",
                            "Spanish Accuracy at 25m",
                            "RT at 25m"))

```



### Partial regression plot: Spanish Composite and Vocabulary
```{r}
# Spanish Composite and Vocab at 25m (from Model 6)

## spanish composite residuals - not including CDI
m_sp_comp <- lm(Spancomposite ~ hi_4.5y_zscore + 
           lbq_4.5y_zscore + rt_zscore, data = transfer)
resid_sp_comp <- resid(m_sp_comp)

# cdi residuals - not including spanish composite
m_cdi_resid <- lm(cdi_25m_zscore ~ hi_4.5y_zscore + 
           lbq_4.5y_zscore + rt_zscore, data = transfer)
resid_cdi <- resid(m_cdi_resid)

# create dataframe
resid_sp_comp_cdi <- data.frame(resid_sp_comp, resid_cdi)

# correlation
corr_sp_comp_cdi <- cor.test(resid_sp_comp, resid_cdi)
round(((corr_sp_comp_cdi$estimate)^2), digits = 3)

# plot
ggplot(resid_sp_comp_cdi, aes(resid_cdi, resid_sp_comp)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  theme(text = element_text(size = 20)) + 
  labs(x = "Spanish Vocabulary at 25m (residuals)", y = "Spanish Language\nComposite\nat 4.5 years\n(residuals)") + 
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5)) + 
  annotate("text", x = 2, y = 26.5, label = "paste(R ^ 2, \" = .169\")", parse = TRUE, size = 6) # change this R2 value for each plot
ggsave("./figures/scatterplot_resid_sp_comp_cdi.pdf", height = 8, width = 11, units = "in")

```



### Partial regression plot: Spanish Composite and RT
```{r}
# Spanish Composite and RT at 25m (from Model 6)

## spanish composite residuals - not including rt
m_sp_comp <- lm(Spancomposite ~ hi_4.5y_zscore + 
           lbq_4.5y_zscore + cdi_25m_zscore, data = transfer)
resid_sp_comp <- resid(m_sp_comp)

# rt residuals - not including spanish composite
m_rt_resid <- lm(rt_zscore ~ hi_4.5y_zscore + 
           lbq_4.5y_zscore + cdi_25m_zscore, data = transfer)
resid_rt <- resid(m_rt_resid)

# create dataframe
resid_sp_comp_rt <- data.frame(resid_sp_comp, resid_rt)

# correlation
corr_sp_comp_rt <- cor.test(resid_sp_comp, resid_rt)
round((corr_sp_comp_rt$estimate)^2, digits = 3)

# plot
ggplot(resid_sp_comp_rt, aes(resid_rt, resid_sp_comp)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  theme(text = element_text(size = 20)) + 
  labs(x = "Spanish RT at 25m (residuals)", y = "Spanish Language\nComposite\nat 4.5 years\n(residuals)") + 
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5)) + 
  annotate("text", x = 2.5, y = 26.5, label = "paste(R ^ 2, \" = .104\")", parse = TRUE, size = 6) # change this R2 value for each plot
ggsave("./figures/scatterplot_resid_sp_comp_rt.pdf", height = 8, width = 11, units = "in")

```



### Partial regression plot: English Composite and Vocabulary
```{r}
# English Composite and Vocab at 25m (from Model 12)

## English composite residuals - not including CDI
m_eng_comp <- lm(Engcomposite ~ hi_4.5y_zscore + 
           lbq_4.5y_zscore + rt_zscore, data = transfer)
resid_eng_comp <- resid(m_eng_comp)

# cdi residuals - not including English composite
m_cdi_resid <- lm(cdi_25m_zscore ~ hi_4.5y_zscore + 
           lbq_4.5y_zscore + rt_zscore, data = transfer)
resid_cdi <- resid(m_cdi_resid)

# create dataframe
resid_eng_comp_cdi <- data.frame(resid_eng_comp, resid_cdi)

# correlation
corr_eng_comp_cdi <- cor.test(resid_eng_comp, resid_cdi)
round(((corr_eng_comp_cdi$estimate)^2), digits = 3)

# plot
ggplot(resid_eng_comp_cdi, aes(resid_cdi, resid_eng_comp)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  theme(text = element_text(size = 20)) + 
  labs(x = "Spanish Vocabulary at 25m (residuals)", y = "English Language\nComposite\nat 4.5 years\n(residuals)") + 
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5)) + 
  annotate("text", x = 2, y = 26.5, label = "paste(R ^ 2, \" = .038\")", parse = TRUE, size = 6) # change this R2 value for each plot
ggsave("./figures/scatterplot_resid_eng_comp_cdi.pdf", height = 8, width = 11, units = "in")

```



### Partial regression plot: English Composite and RT
```{r}
# English Composite and RT at 25m (from Model 12)

## English composite residuals - not including rt
m_eng_comp <- lm(Engcomposite ~ hi_4.5y_zscore + 
           lbq_4.5y_zscore + cdi_25m_zscore, data = transfer)
resid_eng_comp <- resid(m_eng_comp)

# rt residuals - not including English composite
m_rt_resid <- lm(rt_zscore ~ hi_4.5y_zscore + 
           lbq_4.5y_zscore + cdi_25m_zscore, data = transfer)
resid_rt <- resid(m_rt_resid)

# create dataframe
resid_eng_comp_rt <- data.frame(resid_eng_comp, resid_rt)

# correlation
corr_eng_comp_rt <- cor.test(resid_eng_comp, resid_rt)
round((corr_eng_comp_rt$estimate)^2, digits = 3)

# plot
ggplot(resid_eng_comp_rt, aes(resid_rt, resid_eng_comp)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  theme(text = element_text(size = 20)) + 
  labs(x = "Spanish RT at 25m (residuals)", y = "English Language\nComposite\nat 4.5 years\n(residuals)") + 
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5)) + 
  annotate("text", x = 2.5, y = 26.5, label = "paste(R ^ 2, \" = .092\")", parse = TRUE, size = 6) # change this R2 value for each plot
ggsave("./figures/scatterplot_resid_eng_comp_rt.pdf", height = 8, width = 11, units = "in")

```
